Testing traversability of protein tunnels with flexible ligands using motion planning

### 2017-06-07 paper.tex
intro + added images

The widely used tools for tunnel detection are based on Voronoi diagrams~\cite{yaffe2008,caver3} or 
a 3D grid~\cite{sehnal2013mole,petrek2006caver}.
These methods compute the tunnels assuming a spherical ligand (probe).
The resulting tunnels are represented as a sequence of spheres.
The active sites are usually connected with the protein surface with multiple pathways.
However, not all paths are biochemically relevant, so only paths with a given minimal bottleneck have to be reported (the bottleneck is
the radius of the smallest sphere that can traverse the whole tunnel) 
Bio-chemists then decide if such a tunnel can be used to transport a given ligand mainly based on the tunnel lengths and its bottleneck.

@@

Such decision is however not precise and it requires previous expertise with the domain.
Of the reasons is that typical ligands are non-spherical and therefore it is difficult to estimate their traversability in tunnels
computed for a spherical probe.
For better decision, it is necessary to consider also the shape of the ligand being investigated, and consequently, to also
consider rotations of the whole ligand.
Moreover, ligands are rather flexible molecules, so the relative positions of their atoms can change as well.

Sampling-based motion planning methods such as Rapidly Exploring Random Tree~\cite{lavalleRRT} can be
used to test the traversability of a given ligand in the tunnels.
These methods can cope with objects of many degrees of freedom and arbitrary geometry.

@@

\begin{figure}[t]
\centering
{\footnotesize
\renewcommand{\arraystretch}{0.1}
\renewcommand{\tabcolsep}{0pt}
\begin{tabular}{ccc}
\includegraphics[width=0.15\textwidth]{fig/motiv1} &
\includegraphics[width=0.17\textwidth]{fig/motiv2lab} \\
%\includegraphics[width=0.16\textwidth]{fig/motiv3} \\
Protein 1CQW & Active site & Detected tunnels \\
             &            & (orange)
\end{tabular}
}
\caption{\label{fig::motiv}
    Example of tunnel detection in haloalkane dehalogenase.
}
\end{figure}

\red{TODO}
what is traversability analysis.

The goal is the traversability analysis is to estimate traversability of a tunnel with a given ligand from geometrical point of view.

>>>>

The widely used tools for tunnel detection are based on Voronoi diagrams~\cite{yaffe2008,caver3} or 
a 3D grid~\cite{sehnal2013mole,petrek2006caver}.
These methods compute the tunnels assuming a spherical ligand (probe), and the resulting tunnels are represented as a sequence of spheres.
Bio-chemists then decide if such a tunnel can be used to transport a given ligand mainly based on the tunnel lengths and its bottleneck (i.e.,
        the radius of the smallest sphere that can traverse the whole tunnel).
%The active sites are usually connected with the protein surface with multiple pathways.
%However, not all paths are biochemically relevant, so only paths with a given minimal bottleneck have to be reported (the bottleneck is
%the radius of the smallest sphere that can traverse the whole tunnel).
Such decision is however not precise and it requires previous expertise in the domain.

Of the reasons is that typical ligands are non-spherical and therefore it is difficult to estimate their traversability in tunnels
computed for a spherical probe.
For a better decision, it is necessary to consider also the shape of the ligand and possibly also its flexibility.
Due to the increase degrees of freedom (DOF) that need to be considered, the classical Voronoi-based methods cannot be used 
to compute trajectories of the ligands.

@@

In this paper, we propose a novel method for computing the trajectories of the flexible ligands in a tunnel represented by spheres.
The proposed work is based on the sampling-based Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} planner, which
can cope with many-DOF objects of arbitrary geometry.
The random samples for RRT are generated using the spheres of the tunnel.
To make the transportation easier, a scaled-down version of the ligand is used, similarly e.g. to~\cite{cortes2005path}.
The flexiblity of the ligand is modeled using a predefined set of typical conformations.
We propose novel visualization techniques for presenting the results to bio-chemists.
The proposed method can be seen as an extension for the classic tunnel detection tools.

@@

\begin{figure}[t]
\centering
{\footnotesize
\renewcommand{\arraystretch}{0.1}
\renewcommand{\tabcolsep}{0pt}
\begin{tabular}{ccc}
\includegraphics[width=0.15\textwidth]{fig/motiv1} &
\includegraphics[width=0.17\textwidth]{fig/motiv2lab} &
\includegraphics[width=0.16\textwidth]{fig/motiv3} \\
Protein 1CQW & Active site & Detected tunnels \\
             &            & (orange)
\end{tabular}
}
\caption{\label{fig::motiv}
    Example of tunnel detection in haloalkane dehalogenase.
}
\end{figure}
